Large-scale spatiotemporal variation in vital rates of an alpine bird

### 2023-12-06 Manuscript_Nater_etal.qmd
Added flags for Erlend to start drafting
Added text + references about ptarmigan ecology in introduction

-   Shortly about willow ptarmigan (hereafter "ptarmigan"): sentinel species (sensitive to changes in climate and land-use), position in the foodweb (maybe? does it matter?), high cultural value (part of people's "nature experience", popular game species), not redlisted but popular hunting making sustainable management crucial

-   Wrap up with a sentence on why large-scale spatially explicit indicators for ptarmigan populations would be good to have (national and international context)?

@@

>>>>

The willow ptarmigan (*Lagopus lagopus*) has a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian red list of species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. While the main reason for the long-term decline in abundance remain unresolved, they are considered sentinel species that are sensitive to both climate change and land use management [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020]. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can have substantial spatio-temporal variation in demographic rates [@Bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988: @Steen1985]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023, @Steen1985]. Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species. The fact that they are valued game species, information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies. In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is in fact included in both the main national biodiversity (Nature Index for Norway; refs) and ecosystem (Assessment of the Ecological Condition; refs) assessments. The model presented here therefore potentially has direct application in real world management and conservation contexts.

@@

\[End - ERLEND drafts\]
